|

### 2023-08-04 manuscript/paper.Rmd
Manuscript now compiles.  There were some latex errors.

---
title: |
  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$, 
  | \large  Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart${^1,+$} \vspace{-1.1mm}
  |
  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  |
  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  |  
  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  |  
  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  |
  | \large $^*$ Equal contribution \vspace{-1.1mm}
  |
  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
  |
abstract: |
  | Everyone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>). The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reichen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023). The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al. in prep). In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al. 2007). If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G. doryssus</i>.

>>>>

---
title: |
  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$
  | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
  | \vspace{-1.1mm}
  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  | \large $^*$ Equal contribution \vspace{-1.1mm}
  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
abstract: |
  | Everyone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>). The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reimchen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023). The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al. in prep). In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al. 2007). If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G. doryssus</i>.
